---
title: "Reproducibility Report: Group A Choice 1"
author: "Shawn T. Schwartz"
date: "November 3, 2021"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 5
---

-------

For this exercise, please try to reproduce the results from Study 1 of the associated paper (Joel, Teper, & MacDonald, 2014). The PDF of the paper is included in the same folder as this Rmd file. 

#### Methods summary: 

In study 1, 150 introductory psychology students were randomly assigned to a "real" or a "hypothetical" condition. In the real condition, participants believed that they would have a real opportuniy to connect with potential romantic partners. In the hypothetical condition, participants simply imagined that they are on a date. All participants were required to select their favorite profile and answer whether they were willing to exchange contact information. 

------

#### Target outcomes: 

Below is the specific result you will attempt to reproduce (quoted directly from the results section of Study 1):

>> We next tested our primary hypothesis that participants would be more reluctant to reject the unattractive date when they believed the situation to be real rather than hypothetical. Only 10 of the 61 participants in the hypothetical condition chose to exchange contact information with the unattractive potential date (16%). In contrast, 26 of the 71 participants in the real condition chose to exchange contact information (37%). A chi-square test of independence indicated that participants were significantly less likely to reject the unattractive potential date in the real condition compared with the hypothetical condition, X^2(1, N = 132) = 6.77, p = .009.

------

```{r global_options, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, warning=FALSE, message=FALSE)
```

# Step 1: Load packages

```{r}
library(tidyverse) # for data munging
library(knitr) # for kable table formating
library(haven) # import and export 'SPSS', 'Stata' and 'SAS' Files
library(readxl) # import excel files
```

# Step 2: Load data

```{r}
## Just Study 1
d <- read_sav('data/Empathy Gap Study 1 data.sav')
```

# Step 3: Tidy data

```{r}
## confirm participant counts reported in paper
nrow(d) == 132
nrow(subset(d, gender == 0)) == 65 ## male
nrow(subset(d, gender == 1)) == 64 ## female
sum(is.na(d$gender)) == 3 ## unreported

## convert to long format
d_long <- d |>
  select(ID, condition, exchangeinfo) |> ## grab relevant columns
  pivot_longer(cols = c("exchangeinfo"),
               names_to = "Measurement",
               values_to = "Value")
```

# Step 4: Run analysis

## Descriptive statistics

> Only 10 of the 61 participants in the hypothetical condition chose to exchange contact information with the unattractive potential date (16%). In contrast, 26 of the 71 participants in the real condition chose to exchange contact information (37%).

```{r}
## reproduce the above results here

### get data for hypothetical condition
d_hyp_cond <- d_long |>
  filter(condition == 0) |> ## filter out non-hypothetical conditions
  {function(x) table(x$Value)}() |> ## anonymous function to get counts for response value
  {function(x) (x[1] / sum(x))}() |> ## anonymous function to get proportion of yes to total responses
  round(2) ## round response to 2 significant digits

## confirm hypothetical condition results
print(d_hyp_cond)
d_hyp_cond == .16 ## 16%


### get data for real condition
d_real_cond <- d_long |>
  filter(condition == 1) |> ## filter out non-real conditions
  {function(x) table(x$Value)}() |> ## anonymous function to get counts for response value
  {function(x) (x[1] / sum(x))}() |> ## anonymous function to get proportion of yes to total responses
  round(2) ## round response to 2 significant digits

## confirm hypothetical condition results
print(d_real_cond)
d_real_cond == .37 ## 37%
```

## Inferential statistics

> A chi-square test of independence indicated that participants were significantly less likely to reject the unattractive potential date in the real condition compared with the hypothetical condition, X^2(1, N = 132) = 6.77, p = .009.

```{r}
# reproduce the above results here
chisq.test(d_long$condition, d_long$Value, correct = FALSE)
```

# Step 5: Reflection

**Were you able to reproduce the results you attempted to reproduce? If not, what part(s) were you unable to reproduce?**

Yes, I was able to reproduce the results in their entirety.

**How difficult was it to reproduce your results? What aspects made it difficult? What aspects made it easy?**

Not difficult at all, especially since the authors clearly described the columns for the key independent and dependent variables in their dataset. Though, one thing to note was that I had to turn off the Yates' continuity correction for the `chisq.test` function so that the *p*-value and *X-squared* test statistic exactly matched the results reported in the paper.


